\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-20 doc/mlsys_paper/paper.tex
Remove discussion comments from paper for release.
(If we need them back for future revisions, we can just check out the previous
revision and work on a branch or something.)

% TODO: title could possibly be improved
%\title{\texttt{ensmallen}: a generic C++ library for fast optimization}  %% it's not clear what "optimization" refers to here
%% other possibilities:
%\title{\texttt{ensmallen}: a flexible C++ library for function optimization in machine learning}
%\title{\texttt{ensmallen}: a fast C++ library for function optimization in machine learning}
%\title{\texttt{ensmallen}: a C++ library for fast function optimization in machine learning}
%\title{\texttt{ensmallen}: a C++ library for fast and flexible function optimization}
%\title{\texttt{ensmallen}: a C++ library of fast and flexible function optimizers}
%\title{\texttt{ensmallen}: a library of flexible function optimizers in C++}
%\title{\texttt{ensmallen}: a library of fast and flexible function optimizers in C++}
%\title{\texttt{ensmallen}: a flexible C++ library for function optimization}
\title{\texttt{ensmallen}: a flexible C++ library for efficient function optimization}

% Alphabetical ordering?
% TODO: check affiliations
\author{Shikhar Bhardwaj \\
Delhi Technological University \\
Delhi, India 110042 \\
\texttt{shikhar\_bt2k15@dtu.ac.in}
\And
Ryan R. Curtin \\
RelationalAI \\
Atlanta, GA, USA 30318 \\
\texttt{ryan@ratml.org}
\And
Marcus Edel \\
Free University of Berlin \\
Arnimallee 7, 14195 Berlin \\
\texttt{marcus.edel@fu-berlin.de}
\And
%% CS: I've added "Independent Researcher" below for now,
%% CS: as a blank affiliation looks weird and incomplete
Yannis Mentekidis \\
Independent Researcher \\
\texttt{mentekid@gmail.com}
% any affiliation/email?
%% CS: googling suggests that Yannis is/was affiliated with Aristotle University of Thessaloniki 
%% CS: and is perhaps now with Amazon
\And
%% CS: I have two affiliations, so I've listed them on two lines
Conrad Sanderson \\
Data61, CSIRO, Australia \\
University of Queensland, Australia\\
\texttt{conradsand@ieee.org}
}

@@

\begin{abstract}
\vspace*{-0.3em}
%% the abstract below still needs more meat and sharpening
We present \texttt{\small ensmallen}, a fast and flexible C++ library for mathematical optimization of 
arbitrary user-supplied functions,
which can be applied to many machine learning problems.
Several types of optimizations are supported, including differentiable,
separable, constrained, and categorical objective functions.
The library provides many pre-built optimizers
(including numerous variants of SGD and Quasi-Newton optimizers)
as well as a flexible framework for implementing new optimizers and objective functions.
Implementation of a new optimizer requires only one method
and a new objective function requires typically one or two C++ functions. 
This can aid in the quick implementation and prototyping of new machine learning algorithms.
Due to the use of C++ template metaprogramming, \texttt{\small ensmallen} is able to
support compiler optimizations that provide fast runtimes.
Empirical comparisons show that \texttt{\small ensmallen} is able to outperform other
optimization frameworks (like Julia and SciPy), sometimes by large margins.
The library is distributed under the
% save words
% 3-clause
BSD license and is ready for use
in production environments.
\end{abstract}

@@

Mathematical optimization is the workhorse of virtually all machine learning
algorithms.  For a given objective function $f(\cdot)$
(which may have a special structure or constraints),
almost all machine learning problems can be boiled down
to the following optimization form:
%
%\vspace*{-0.2em}
\begin{equation}
\argmindown_x f(x).
\end{equation}
\vspace*{-1.5em}

Optimization is often computationally intensive and may correspond
to most of the time taken to train a machine learning model.  For instance, the
training of deep neural networks is dominated by the optimization of the model
parameters on the data~\cite{schmidhuber2015deep}.
Even popular machine learning models such as logistic regression
have training times mostly dominated by an optimization procedure~\cite{kingma2015adam}.
% TODO: might be nice to have something kind of anecdotal like 'even new
% students to the field of machine learning quickly encounter optimization' and
% cite, e.g., Andrew Ng's coursera course or some ML textbook or similar
%% CS: i think we don't need to explore this too much; better to cut out all the
%% CS: fat and stick with concrete examples, instead of veering off on tangents
%
% or maybe just a note about how many optimization techniques get published at
% NIPS every year?
%% CS: NIPS is too self-referential here

@@

%   \begin{tablenotes}\footnotesize

@@

\texttt{\small ensmallen} provides a {\bf set of optimizers} for
optimizing {\bf user-defined objective functions}.  It is also easy to implement a
new optimizer in the \texttt{\small ensmallen} framework.  Overall, our goal is to provide
an easy-to-use library that can solve the problem
%\vspace*{-0.4em}
%\begin{equation}
$\argminright_{x} f(x)$
%\end{equation}
%\vspace*{-0.4em}
%\noindent
for any function $f(x)$ that takes a vector or matrix input $x$.
In most cases, $f(x)$ will have special structure; one example might be that
$f(x)$ is differentiable.  Therefore, the abstraction we have designed for \texttt{\small
ensmallen} can optionally take advantage of this structure.  For example, in
addition to $f(x)$, a user can provide an implementation of $f'(x)$, which in
turn allows first-order gradient-based optimizers to be used.  This generally
leads to significant speedups.

@@

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\bf arbitrary}: no assumptions can be made on $f(x)$
  \item {\bf differentiable}: $f(x)$ has a computable gradient $f'(x)$
  \item {\bf separable}: $f(x)$ is a sum of individual components: $f(x) =
\sum_{i} f_i(x)$
  \item {\bf categorical}: $x$ contains elements that can only take discrete
values
  %\item {\bf numeric}: all elements of $x$ take values in $\mathcal{R}$
  \item {\bf sparse}: the gradient $f'(x)$ or $f'_i(x)$ (for a separable
function) is sparse
  \item {\bf partially differentiable}: the separable gradient $f_i'(x)$ is also
separable for a different axis $j$
  \item {\bf bounded}: $x$ is limited in the values that it can take
\end{itemize}
\vspace*{-0.3em}

@@

%% CS: WARNING !!!!
%% CS: can't add more citations without causing the item with SGD variants
%% CS: to overflow into 3 lines.
%% CS: this causes a cascade effect of mucking up the entire layout of
%% CS: the paper, causing the main text to spill over to 7 pages.
%% 
%% CS: the citations below should be sufficient; 
%% CS: this is a workshop paper, not a journal article

@@

  \item {\bf Other}:
    Conditional Gradient Descent,
    Frank-Wolfe algorithm~\cite{Frank_1956},
    Simulated Annealing~\cite{kirkpatrick1983optimization}

% These were a part of mlpack but not ensmallen.
  %\item {\bf Objective functions:} Neural Networks, Logistic regression,
  %    Matrix completion, Neighborhood Components Analysis, Regularized SVD,
  %    Reinforcement learning, Softmax regression, Sparse autoencoders,
  %    Sparse SVM
\end{enumerate}

@@

%% CS: i've added \nolimits to save a bit of space
\vspace*{-0.5em}
\begin{equation}
\argmindown_\theta f(\theta) =   %% CS: for clarity
\argmindown_\theta \sum\nolimits_{i = 1}^n (y_i - x_i \theta)^2 =
\argmindown_\theta \| y - X \theta \|_F^2.
\end{equation}
\vspace*{-0.5em}

@@

%\vspace*{-0.5em}
\begin{adjustbox}{scale={0.95}{0.95}}
\begin{minipage}{1\textwidth}
\begin{minted}[fontsize=\small]{c++}
    double EvaluateWithGradient(const arma::mat& theta, arma::mat& gradient) {
      const arma::vec v = (y - X * theta); // Cache result
      gradient = -2 * X.t() * v;  // Store gradient in the provided matrix
      return arma::accu(v % v); // Take squared norm of v
    }
\end{minted}
\end{minipage}
\end{adjustbox}
%\vspace*{-0.5em}

@@

% TODO: anything to write about the visualization page that we had set up?

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{lcccc}
\toprule
 & \texttt{\small ensmallen} & \texttt{\small scipy} & \texttt{\small Optim.jl} & \texttt{\small samin} \\
\midrule
% TODO: these are just single-run results from Marcus' laptop!  We need to do
% 10 and average.
default & {\bf 0.004s} & 1.069s & 0.021s & 3.173s \\
tuned & & 0.574s & & 3.122s \\ % TODO
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for $100000$ iterations of simulated annealing with various
frameworks on the simple Rosenbrock function.  Julia code runs do not count
compilation time.  The {\it tuned} row indicates that the code was manually
modified for speed.}
\label{tab:rosenbrock_results}
\end{table}

@@

We compare four frameworks for this task:
{\bf (i)}~\texttt{\small ensmallen},
{\bf (ii)}~\texttt{\small scipy.optimize.anneal} from SciPy 0.14.1~\cite{jones2014scipy},
{\bf (iii)}~simulated annealing implementation in \texttt{\small Optim.jl} with Julia 1.0.1~\cite{mogensen2018optim},
{\bf (iv)}~\texttt{\small samin} in the \texttt{\small optim} package for Octave~\cite{octave}.
While another option here might be \texttt{\small simulannealbnd()} 
in the Global Optimization Toolkit for MATLAB,
no license was available.
% TODO: get Marcus' system specs.
We ran our code on a MacBook Pro i7 2018 with 16GB RAM running macOS 10.14 with clang 1000.10.44.2, Julia version 1.0.1, Python 2.7.15, and Octave 4.4.1.

% We compare four frameworks%
% %
% \footnote{Another option here might be \texttt{\small simulannealbnd()} 
% in the Global Optimization Toolkit for MATLAB.
% However, no license was available for these simulations.}
% %
% for this task:
% 
% \vspace*{-0.3em}
% \begin{itemize}
% \renewcommand{\itemsep}{-0.5ex}
%   \item \texttt{\small ensmallen}
%   \item \texttt{\small scipy.optimize.anneal}, from scipy 0.14.1~\cite{jones2014scipy}
%   \item simulated annealing implementation in \texttt{\small Optim.jl} with Julia
% 1.0.1~\cite{mogensen2018optim}
%   \item \texttt{\small samin} in the \texttt{\small optim} package for GNU Octave~\cite{octave}
% \end{itemize}
% \vspace*{-0.3em}

@@

% For \texttt{\small ensmallen} we have 2 versions:
% (i)~with only \texttt{\small EvaluateWithGradient()},
% and
% (ii)~with \texttt{\small Evaluate()} and \texttt{\small Gradient()}.
% The code for these functions is as shown earlier.
% For Julia we have the options of using manually defined objective and gradient functions,
% or the gradient function can be automatically computed by 
% \texttt{\small Calculus.jl}
% (\href{https://github.com/JuliaMath/Calculus.jl}{\footnotesize github.com/JuliaMath/Calculus.jl})
% or \texttt{\small ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016}.

@@

\begin{table}
\centering
\begin{adjustbox}{scale={0.90}{0.90}}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d$: 100, $n$: 1k & $d$: 100, $n$: 10k & $d$: 100, $n$:
100k & $d$: 1k, $n$: 100k \\
\midrule
% TODO: this was only one trial on Ryan's desktop!
\texttt{\small ensmallen}-1 & {\bf 0.001s} & {\bf 0.009s} & {\bf 0.154s} & {\bf 2.215s} \\
\texttt{\small ensmallen}-2 & 0.002s & 0.016s & 0.182s & 2.522s \\
% Dropped for space and awful performance
%\texttt{\small Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
\texttt{\small Optim.jl} & 0.006s & 0.030s & 0.337s & 4.271s \\
\texttt{\small scipy} & 0.003s & 0.017s & 0.202s & 2.729s \\
\texttt{\small bfgsmin} & 0.071s & 0.859s & 23.220s & 2859.81s\\
% It's possible to tune ForwardDiff.jl a bit, but it doesn't give significant
% speedups to make it competitive and it really makes the code ugly.
\texttt{\small ForwardDiff.jl} & 0.497s & 1.159s & 4.996s & 603.106s \\
\texttt{\small autograd} & 0.007s & 0.026s & 0.210s & 2.673s \\
\bottomrule
\end{tabular}
\end{adjustbox}
\vspace*{0.25ex}
\caption{\footnotesize
Runtimes for the linear regression function on various dataset sizes,
with $n$ indicating the number of samples,
and $d$ indicating the dimensionality of each sample.
All Julia runs do not count compilation time.}
\label{tab:lbfgs}
%\vspace*{-1ex}
\end{table}

@@

% TODO: show flexibility of optimization with learning curves:
%  - use LinearRegressionFunction modified for small batches
%  - make sure Info or Debug output is on
%  - run with a whole boatload of SGD variants
%  - parse the output with awk/sed into a csv of objectives per epoch
%  - plot it
%  - profit!
%
% Probably a snippet showing the actual code to run with a bunch of different
% optimizers is good too.  Other things can be cut to make space.

Lastly, we demonstrate the easy pluggability in \texttt{\small ensmallen}
for using various optimizers on the same task.
Using a version of \texttt{\small LinearRegressionFunction} from Sec.~\ref{sec:linreg_example}
adapted for separable differentiable optimizers,
we run six optimizers with default parameters in just 8 lines of code,
as shown in Fig.~\ref{fig:learning_curve}(a).
Applying these optimizers to the \texttt{\small YearPredictionMSD}
dataset from the UCI repository~\cite{ucimlrepository}
yields the learning curves shown in Fig.~\ref{fig:learning_curve}(b).
Any other optimizer for separable differentiable objective
functions can be dropped into place in the same manner.
%% Just because we have some extra space...
%% CS: we need space for the acknowledgement section
% This facilitates the seamless evaluation of various optimizers
% for user-defined objective functions.

@@

% RC: I think it's really important to highlight ensmallen's usage (and
% genesis), although I can't find the right words to concisely and non-awkwardly
% say that we wrote ensmallen as part of mlpack originally.
% 
%% CS: good point, though for our purposes i think it's sufficient
%% CS: to simply state that mlpack uses ensmallen.
%% CS: getting into a tangent on the genesis can negatively distract
%% CS: from the central message. besides, there is no room for a proper
%% CS: explanation.
%% CS:
%% CS: I recommend to avoid interchangeably mixing around the words
%% CS: "library", "toolkit", "package" when referring to ensmallen.
%% CS: it's better to consistently stick to "library", and use
%% CS: the other words to refer to other software, such as mlpack.
%% CS: the point is to avoid potential concept clashes
%% CS: (ie. too much overloading on a word), which can lead
%% CS: to confusion as to what exact software we're referring to.

%\begin{small}
{\bf Acknowledgements.}
We would like to thank the many contributors to \texttt{\small ensmallen},
who are listed on the associated website.
%\end{small}

% \subsubsection*{Acknowledgements}
% \vspace*{-0.5em}
% 
% The development team of \texttt{\small ensmallen} does not include just the authors named
% here but also a long list of other contributors.  See
% \url{https://www.ensmallen.org/about.html} for more information.
% % TODO: that URL may change

>>>>

\title{\texttt{ensmallen}: a flexible C++ library for efficient function optimization}

\author{Shikhar Bhardwaj \\
Delhi Technological University \\
Delhi, India 110042 \\
\texttt{shikhar\_bt2k15@dtu.ac.in}
\And
Ryan R. Curtin \\
RelationalAI \\
Atlanta, GA, USA 30318 \\
\texttt{ryan@ratml.org}
\And
Marcus Edel \\
Free University of Berlin \\
Arnimallee 7, 14195 Berlin \\
\texttt{marcus.edel@fu-berlin.de}
\And
Yannis Mentekidis \\
Independent Researcher \\
\texttt{mentekid@gmail.com}
\And
Conrad Sanderson \\
Data61, CSIRO, Australia \\
University of Queensland, Australia\\
\texttt{conradsand@ieee.org}
}

@@

\begin{abstract}
\vspace*{-0.3em}
We present \texttt{\small ensmallen}, a fast and flexible C++ library for mathematical optimization of 
arbitrary user-supplied functions,
which can be applied to many machine learning problems.
Several types of optimizations are supported, including differentiable,
separable, constrained, and categorical objective functions.
The library provides many pre-built optimizers
(including numerous variants of SGD and Quasi-Newton optimizers)
as well as a flexible framework for implementing new optimizers and objective functions.
Implementation of a new optimizer requires only one method
and a new objective function requires typically one or two C++ functions. 
This can aid in the quick implementation and prototyping of new machine learning algorithms.
Due to the use of C++ template metaprogramming, \texttt{\small ensmallen} is able to
support compiler optimizations that provide fast runtimes.
Empirical comparisons show that \texttt{\small ensmallen} is able to outperform other
optimization frameworks (like Julia and SciPy), sometimes by large margins.
The library is distributed under the
BSD license and is ready for use
in production environments.
\end{abstract}

@@

Mathematical optimization is the workhorse of virtually all machine learning
algorithms.  For a given objective function $f(\cdot)$
(which may have a special structure or constraints),
almost all machine learning problems can be boiled down
to the following optimization form:

\begin{equation}
\argmindown_x f(x).
\end{equation}
\vspace*{-1.5em}

Optimization is often computationally intensive and may correspond
to most of the time taken to train a machine learning model.  For instance, the
training of deep neural networks is dominated by the optimization of the model
parameters on the data~\cite{schmidhuber2015deep}.
Even popular machine learning models such as logistic regression
have training times mostly dominated by an optimization procedure~\cite{kingma2015adam}.

@@

@@

\texttt{\small ensmallen} provides a {\bf set of optimizers} for
optimizing {\bf user-defined objective functions}.  It is also easy to implement a
new optimizer in the \texttt{\small ensmallen} framework.  Overall, our goal is to provide
an easy-to-use library that can solve the problem
$\argminright_{x} f(x)$
for any function $f(x)$ that takes a vector or matrix input $x$.
In most cases, $f(x)$ will have special structure; one example might be that
$f(x)$ is differentiable.  Therefore, the abstraction we have designed for \texttt{\small
ensmallen} can optionally take advantage of this structure.  For example, in
addition to $f(x)$, a user can provide an implementation of $f'(x)$, which in
turn allows first-order gradient-based optimizers to be used.  This generally
leads to significant speedups.

@@

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\bf arbitrary}: no assumptions can be made on $f(x)$
  \item {\bf differentiable}: $f(x)$ has a computable gradient $f'(x)$
  \item {\bf separable}: $f(x)$ is a sum of individual components: $f(x) =
\sum_{i} f_i(x)$
  \item {\bf categorical}: $x$ contains elements that can only take discrete
values
  \item {\bf sparse}: the gradient $f'(x)$ or $f'_i(x)$ (for a separable
function) is sparse
  \item {\bf partially differentiable}: the separable gradient $f_i'(x)$ is also
separable for a different axis $j$
  \item {\bf bounded}: $x$ is limited in the values that it can take
\end{itemize}
\vspace*{-0.3em}

@@

@@

  \item {\bf Other}:
    Conditional Gradient Descent,
    Frank-Wolfe algorithm~\cite{Frank_1956},
    Simulated Annealing~\cite{kirkpatrick1983optimization}
\end{enumerate}

@@

\vspace*{-0.5em}
\begin{equation}
\argmindown_\theta f(\theta) =
\argmindown_\theta \sum\nolimits_{i = 1}^n (y_i - x_i \theta)^2 =
\argmindown_\theta \| y - X \theta \|_F^2.
\end{equation}
\vspace*{-0.5em}

@@

\begin{adjustbox}{scale={0.95}{0.95}}
\begin{minipage}{1\textwidth}
\begin{minted}[fontsize=\small]{c++}
    double EvaluateWithGradient(const arma::mat& theta, arma::mat& gradient) {
      const arma::vec v = (y - X * theta); // Cache result
      gradient = -2 * X.t() * v;  // Store gradient in the provided matrix
      return arma::accu(v % v); // Take squared norm of v
    }
\end{minted}
\end{minipage}
\end{adjustbox}

@@

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{lcccc}
\toprule
 & \texttt{\small ensmallen} & \texttt{\small scipy} & \texttt{\small Optim.jl} & \texttt{\small samin} \\
\midrule
default & {\bf 0.004s} & 1.069s & 0.021s & 3.173s \\
tuned & & 0.574s & & 3.122s \\
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for $100000$ iterations of simulated annealing with various
frameworks on the simple Rosenbrock function.  Julia code runs do not count
compilation time.  The {\it tuned} row indicates that the code was manually
modified for speed.}
\label{tab:rosenbrock_results}
\end{table}

@@

We compare four frameworks for this task:
{\bf (i)}~\texttt{\small ensmallen},
{\bf (ii)}~\texttt{\small scipy.optimize.anneal} from SciPy 0.14.1~\cite{jones2014scipy},
{\bf (iii)}~simulated annealing implementation in \texttt{\small Optim.jl} with Julia 1.0.1~\cite{mogensen2018optim},
{\bf (iv)}~\texttt{\small samin} in the \texttt{\small optim} package for Octave~\cite{octave}.
While another option here might be \texttt{\small simulannealbnd()} 
in the Global Optimization Toolkit for MATLAB,
no license was available.
We ran our code on a MacBook Pro i7 2018 with 16GB RAM running macOS 10.14 with clang 1000.10.44.2, Julia version 1.0.1, Python 2.7.15, and Octave 4.4.1.

@@

@@

\begin{table}
\centering
\begin{adjustbox}{scale={0.90}{0.90}}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d$: 100, $n$: 1k & $d$: 100, $n$: 10k & $d$: 100, $n$:
100k & $d$: 1k, $n$: 100k \\
\midrule
\texttt{\small ensmallen}-1 & {\bf 0.001s} & {\bf 0.009s} & {\bf 0.154s} & {\bf 2.215s} \\
\texttt{\small ensmallen}-2 & 0.002s & 0.016s & 0.182s & 2.522s \\
% Dropped for space and awful performance
%\texttt{\small Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
\texttt{\small Optim.jl} & 0.006s & 0.030s & 0.337s & 4.271s \\
\texttt{\small scipy} & 0.003s & 0.017s & 0.202s & 2.729s \\
\texttt{\small bfgsmin} & 0.071s & 0.859s & 23.220s & 2859.81s\\
% It's possible to tune ForwardDiff.jl a bit, but it doesn't give significant
% speedups to make it competitive and it really makes the code ugly.
\texttt{\small ForwardDiff.jl} & 0.497s & 1.159s & 4.996s & 603.106s \\
\texttt{\small autograd} & 0.007s & 0.026s & 0.210s & 2.673s \\
\bottomrule
\end{tabular}
\end{adjustbox}
\vspace*{0.25ex}
\caption{\footnotesize
Runtimes for the linear regression function on various dataset sizes,
with $n$ indicating the number of samples,
and $d$ indicating the dimensionality of each sample.
All Julia runs do not count compilation time.}
\label{tab:lbfgs}
\end{table}

@@

Lastly, we demonstrate the easy pluggability in \texttt{\small ensmallen}
for using various optimizers on the same task.
Using a version of \texttt{\small LinearRegressionFunction} from Sec.~\ref{sec:linreg_example}
adapted for separable differentiable optimizers,
we run six optimizers with default parameters in just 8 lines of code,
as shown in Fig.~\ref{fig:learning_curve}(a).
Applying these optimizers to the \texttt{\small YearPredictionMSD}
dataset from the UCI repository~\cite{ucimlrepository}
yields the learning curves shown in Fig.~\ref{fig:learning_curve}(b).
Any other optimizer for separable differentiable objective
functions can be dropped into place in the same manner.

@@

{\bf Acknowledgements.}
We would like to thank the many contributors to \texttt{\small ensmallen},
who are listed on the associated website.
